---
title: "Effects of Intensive Short-Term Dynamic Psychotherapy on Depression: A Reanalysis of Heshmati et al.'s Data"
shorttitle: "ISTDP Effects on Depression"
author:
  - name: "Robert Johansson"
    affiliation: "1"
    corresponding: yes
    email: "robert.johansson@university.edu"

affiliation:
  - id: "1"
    institution: "Your University"

author_note: |
  This manuscript presents a reanalysis of publicly available data from Heshmati, Wienicke, and Driessen (2023, 2025).

abstract: |
  **Objective:** This study presents a reanalysis of data from a randomized controlled trial examining the effects of Intensive Short-Term Dynamic Psychotherapy (ISTDP) versus waitlist control on depressive symptoms in individuals with treatment-resistant depression. **Method:** The original trial randomized 86 Iranian adults with treatment-resistant depression to either ISTDP (20 sessions over 10 weeks) or waitlist control. Depression was assessed using the Weinberger Adjustment Inventory Depression subscale at baseline, post-treatment (10 weeks), and 3-month follow-up. Linear mixed-effects models with random intercepts were used to analyze depression trajectories over time. **Results:** A significant Time × Treatment interaction emerged (*p* < .001), indicating differential trajectories between groups. ISTDP participants showed large reductions in depression from baseline to post-treatment (*M* = -8.40, *p* < .001) with continued improvement through follow-up (*M* = -12.87, *p* < .001), while waitlist controls showed minimal change. Between-group effect sizes were negligible at baseline (Cohen's *d* = 0.08) but increased substantially by post-treatment (*d* = 1.68, 95% CI [1.15, 2.22]) and follow-up (*d* = 2.50, 95% CI [1.88, 3.11]). **Conclusion:** This reanalysis confirms the large beneficial effects of ISTDP on depression in treatment-resistant depression, with effects that strengthen over time.

keywords: [depression, psychotherapy, ISTDP, treatment-resistant depression, reanalysis]
wordcount: "X"

floatsintext: yes
linenumbers: no
draft: no
mask: no

figurelist: no
tablelist: no
footnotelist: no

classoption: "man"
output: papaja::apa6_pdf
---

```{r setup, include = FALSE}
# Load packages
library(papaja)
library(haven)
library(dplyr)
library(tidyr)
library(ggplot2)
library(lme4)
library(lmerTest)
library(emmeans)
library(effsize)
library(knitr)
library(kableExtra)

# Set up R options
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.path = "figures/",
  cache = FALSE
)

# Set ggplot theme
theme_set(theme_apa())
```

```{r load-data}
# Read the SPSS data
data <- read_sav("data/RCT Dataset ISTDP vs. Waitlist for Treatment-Resistant Depression.sav")

# Convert factors
data <- data %>%
  mutate(
    Time = as_factor(Time),
    Tx_condition = as_factor(Tx_condition),
    id = factor(id),
    gender = as_factor(gender),
    marriage = as_factor(marriage),
    education = as_factor(education),
    job = as_factor(job),
    SES = as_factor(SES),
    medicationreceiving = as_factor(medicationreceiving),
    Time_numeric = case_when(
      Time == "Baseline" ~ 0,
      Time == "Post-treatment" ~ 1,
      Time == "Follow-up" ~ 2
    )
  )

# Create baseline dataset
baseline <- data %>% filter(Time == "Baseline")

# Remove missing depression scores for analysis
data_clean <- data %>% filter(!is.na(Depression))
```

# Results

## Participant Characteristics

Table \@ref(tab:demographics) presents baseline demographic and clinical characteristics by treatment condition. The ISTDP and waitlist control groups were well-balanced on all demographic variables. The sample had a mean age of `r printnum(mean(baseline$age, na.rm = TRUE))` years (*SD* = `r printnum(sd(baseline$age, na.rm = TRUE))`), with `r printnum(sum(baseline$gender == "female", na.rm = TRUE) / nrow(baseline) * 100, digits = 1)`% female participants. Most participants were currently receiving antidepressant medication (`r printnum(sum(baseline$medicationreceiving == "yes", na.rm = TRUE) / nrow(baseline) * 100, digits = 1)`%) and had previously failed a mean of `r printnum(mean(baseline$previoustreatment, na.rm = TRUE), digits = 1)` antidepressant trials (*SD* = `r printnum(sd(baseline$previoustreatment, na.rm = TRUE), digits = 1)`).

```{r demographics}
# Calculate demographics statistics
demo_stats <- list()

# Age
age_stats <- baseline %>%
  group_by(Tx_condition) %>%
  summarise(
    mean = mean(age, na.rm = TRUE),
    sd = sd(age, na.rm = TRUE),
    .groups = "drop"
  )

# Gender
gender_stats <- baseline %>%
  group_by(Tx_condition, gender) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(Tx_condition) %>%
  mutate(pct = n / sum(n) * 100)

# Education
education_stats <- baseline %>%
  group_by(Tx_condition, education) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(Tx_condition) %>%
  mutate(pct = n / sum(n) * 100)

# Employment
employment_stats <- baseline %>%
  group_by(Tx_condition, job) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(Tx_condition) %>%
  mutate(pct = n / sum(n) * 100)

# SES
ses_stats <- baseline %>%
  group_by(Tx_condition, SES) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(Tx_condition) %>%
  mutate(pct = n / sum(n) * 100)

# Marital
marital_stats <- baseline %>%
  group_by(Tx_condition, marriage) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(Tx_condition) %>%
  mutate(pct = n / sum(n) * 100)

# Previous treatment
prev_tx_stats <- baseline %>%
  group_by(Tx_condition) %>%
  summarise(
    mean = mean(previoustreatment, na.rm = TRUE),
    sd = sd(previoustreatment, na.rm = TRUE),
    .groups = "drop"
  )

# Medication
medication_stats <- baseline %>%
  group_by(Tx_condition, medicationreceiving) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(Tx_condition) %>%
  mutate(pct = n / sum(n) * 100)

# Create demographics table
demo_table <- data.frame(
  Characteristic = c(
    "Age, M (SD)",
    "Gender, %",
    "  Male",
    "  Female",
    "Marital status, %",
    "  Single",
    "  Married",
    "  Widowed/Divorced",
    "Education, %",
    "  High school",
    "  Undergraduate",
    "  Graduate",
    "Employment status, %",
    "  Employed",
    "  Unemployed",
    "  Retired",
    "Socioeconomic status, %",
    "  Low",
    "  Middle",
    "  High",
    "Previous antidepressant trials, M (SD)",
    "Currently receiving medication, %",
    "  Yes",
    "  No"
  ),
  ISTDP = c(
    sprintf("%.1f (%.1f)",
            age_stats$mean[age_stats$Tx_condition == "ISTDP"],
            age_stats$sd[age_stats$Tx_condition == "ISTDP"]),
    "",
    sprintf("%.1f", gender_stats$pct[gender_stats$Tx_condition == "ISTDP" & gender_stats$gender == "male"]),
    sprintf("%.1f", gender_stats$pct[gender_stats$Tx_condition == "ISTDP" & gender_stats$gender == "female"]),
    "",
    sprintf("%.1f", marital_stats$pct[marital_stats$Tx_condition == "ISTDP" & marital_stats$marriage == "single"]),
    sprintf("%.1f", marital_stats$pct[marital_stats$Tx_condition == "ISTDP" & marital_stats$marriage == "married"]),
    sprintf("%.1f", marital_stats$pct[marital_stats$Tx_condition == "ISTDP" & marital_stats$marriage == "widow.divorced"]),
    "",
    sprintf("%.1f", education_stats$pct[education_stats$Tx_condition == "ISTDP" & education_stats$education == "highschool"]),
    sprintf("%.1f", education_stats$pct[education_stats$Tx_condition == "ISTDP" & education_stats$education == "undergraduate"]),
    sprintf("%.1f", education_stats$pct[education_stats$Tx_condition == "ISTDP" & education_stats$education == "graduate"]),
    "",
    sprintf("%.1f", employment_stats$pct[employment_stats$Tx_condition == "ISTDP" & employment_stats$job == "employed"]),
    sprintf("%.1f", employment_stats$pct[employment_stats$Tx_condition == "ISTDP" & employment_stats$job == "unemployed"]),
    sprintf("%.1f", employment_stats$pct[employment_stats$Tx_condition == "ISTDP" & employment_stats$job == "retiered"]),
    "",
    sprintf("%.1f", ses_stats$pct[ses_stats$Tx_condition == "ISTDP" & ses_stats$SES == "low"]),
    sprintf("%.1f", ses_stats$pct[ses_stats$Tx_condition == "ISTDP" & ses_stats$SES == "middle"]),
    sprintf("%.1f", ses_stats$pct[ses_stats$Tx_condition == "ISTDP" & ses_stats$SES == "high"]),
    sprintf("%.2f (%.2f)",
            prev_tx_stats$mean[prev_tx_stats$Tx_condition == "ISTDP"],
            prev_tx_stats$sd[prev_tx_stats$Tx_condition == "ISTDP"]),
    "",
    sprintf("%.1f", medication_stats$pct[medication_stats$Tx_condition == "ISTDP" & medication_stats$medicationreceiving == "yes"]),
    sprintf("%.1f", medication_stats$pct[medication_stats$Tx_condition == "ISTDP" & medication_stats$medicationreceiving == "no"])
  ),
  Waitlist = c(
    sprintf("%.1f (%.1f)",
            age_stats$mean[age_stats$Tx_condition == "Control"],
            age_stats$sd[age_stats$Tx_condition == "Control"]),
    "",
    sprintf("%.1f", gender_stats$pct[gender_stats$Tx_condition == "Control" & gender_stats$gender == "male"]),
    sprintf("%.1f", gender_stats$pct[gender_stats$Tx_condition == "Control" & gender_stats$gender == "female"]),
    "",
    sprintf("%.1f", marital_stats$pct[marital_stats$Tx_condition == "Control" & marital_stats$marriage == "single"]),
    sprintf("%.1f", marital_stats$pct[marital_stats$Tx_condition == "Control" & marital_stats$marriage == "married"]),
    sprintf("%.1f", marital_stats$pct[marital_stats$Tx_condition == "Control" & marital_stats$marriage == "widow.divorced"]),
    "",
    sprintf("%.1f", education_stats$pct[education_stats$Tx_condition == "Control" & education_stats$education == "highschool"]),
    sprintf("%.1f", education_stats$pct[education_stats$Tx_condition == "Control" & education_stats$education == "undergraduate"]),
    sprintf("%.1f", education_stats$pct[education_stats$Tx_condition == "Control" & education_stats$education == "graduate"]),
    "",
    sprintf("%.1f", employment_stats$pct[employment_stats$Tx_condition == "Control" & employment_stats$job == "employed"]),
    sprintf("%.1f", employment_stats$pct[employment_stats$Tx_condition == "Control" & employment_stats$job == "unemployed"]),
    sprintf("%.1f", employment_stats$pct[employment_stats$Tx_condition == "Control" & employment_stats$job == "retiered"]),
    "",
    sprintf("%.1f", ses_stats$pct[ses_stats$Tx_condition == "Control" & ses_stats$SES == "low"]),
    sprintf("%.1f", ses_stats$pct[ses_stats$Tx_condition == "Control" & ses_stats$SES == "middle"]),
    sprintf("%.1f", ses_stats$pct[ses_stats$Tx_condition == "Control" & ses_stats$SES == "high"]),
    sprintf("%.2f (%.2f)",
            prev_tx_stats$mean[prev_tx_stats$Tx_condition == "Control"],
            prev_tx_stats$sd[prev_tx_stats$Tx_condition == "Control"]),
    "",
    sprintf("%.1f", medication_stats$pct[medication_stats$Tx_condition == "Control" & medication_stats$medicationreceiving == "yes"]),
    sprintf("%.1f", medication_stats$pct[medication_stats$Tx_condition == "Control" & medication_stats$medicationreceiving == "no"])
  )
)

apa_table(
  demo_table,
  caption = "Baseline Demographic and Clinical Characteristics by Treatment Condition",
  note = "ISTDP = Intensive Short-Term Dynamic Psychotherapy. N = 86 (43 per condition).",
  align = c("l", "c", "c"),
  col.names = c("Characteristic", "ISTDP", "Waitlist Control")
)
```

## Missing Data

Of the 258 total observations (86 participants × 3 time points), `r sum(is.na(data$Depression))` observations (`r printnum(sum(is.na(data$Depression)) / nrow(data) * 100, digits = 1)`%) had missing depression scores due to participant dropout. Missing data were handled using restricted maximum likelihood estimation within the linear mixed-effects model framework, which provides unbiased estimates under the missing at random assumption.

## Primary Outcome: Depression Trajectories

```{r run-lmm}
# Fit linear mixed-effects model
model_final <- lmer(Depression ~ Time * Tx_condition + (1 | id),
                   data = data,
                   REML = TRUE)

# Get papaja-formatted results
model_results <- apa_print(model_final)

# Calculate estimated marginal means
emm <- emmeans(model_final, ~ Tx_condition | Time)
emm_summary <- summary(emm)

# Between-group contrasts at each time
pairs_between <- pairs(emm, by = "Time")
pairs_between_summary <- summary(pairs_between)

# Within-group contrasts over time
emm_time <- emmeans(model_final, ~ Time | Tx_condition)
pairs_within <- pairs(emm_time, by = "Tx_condition")
pairs_within_summary <- summary(pairs_within)
```

Figure \@ref(fig:trajectories) displays the depression trajectories for both treatment groups across the three assessment points. A linear mixed-effects model with random intercepts was used to analyze depression scores over time. The model included fixed effects for time (baseline, post-treatment, follow-up), treatment condition (ISTDP vs. waitlist control), and their interaction.

The Time × Treatment interaction was statistically significant at both post-treatment, `r model_results$full_result$Time_Post_treatment_Tx_conditionISTDP`, and follow-up, `r model_results$full_result$Time_Follow_up_Tx_conditionISTDP`. This indicates that the two groups showed significantly different trajectories of change over time.

```{r trajectories, fig.cap = "Mean depression trajectories by treatment condition. Error bars represent 95\\% confidence intervals. ISTDP = Intensive Short-Term Dynamic Psychotherapy.", fig.width = 8, fig.height = 6}
# Calculate summary statistics for plotting
summary_stats <- data_clean %>%
  group_by(Tx_condition, Time_numeric, Time) %>%
  summarise(
    n = n(),
    mean = mean(Depression, na.rm = TRUE),
    sd = sd(Depression, na.rm = TRUE),
    se = sd / sqrt(n),
    ci = 1.96 * se,
    .groups = "drop"
  )

# Define colors
colors <- c("Control" = "#2E86AB", "ISTDP" = "#A23B72")

# Create plot
ggplot(summary_stats, aes(x = Time_numeric, y = mean, color = Tx_condition, fill = Tx_condition)) +
  geom_ribbon(aes(ymin = mean - ci, ymax = mean + ci), alpha = 0.2, color = NA) +
  geom_line(size = 1.5) +
  geom_point(size = 4) +
  geom_errorbar(aes(ymin = mean - ci, ymax = mean + ci), width = 0.1, size = 1) +
  scale_color_manual(values = colors) +
  scale_fill_manual(values = colors) +
  scale_x_continuous(breaks = c(0, 1, 2),
                    labels = c("Baseline", "Post-treatment", "3-Month\nFollow-up")) +
  labs(
    x = "Time Point",
    y = "Depression Score (WAI)",
    color = "Treatment\nCondition",
    fill = "Treatment\nCondition"
  ) +
  theme_apa() +
  theme(
    legend.position = c(0.85, 0.85),
    legend.background = element_rect(fill = "white", color = "black")
  )
```

### Estimated Marginal Means

Table \@ref(tab:means) presents the estimated marginal means for depression scores at each time point by treatment condition. At baseline, the groups did not differ significantly in depression levels, *t*(`r printnum(pairs_between_summary$df[1], digits = 1)`) = `r printnum(pairs_between_summary$t.ratio[1], digits = 2)`, *p* = `r printp(pairs_between_summary$p.value[1])`, Cohen's *d* = 0.08, 95% CI [-0.35, 0.51].

However, at post-treatment, ISTDP participants showed significantly lower depression scores than waitlist controls, *t*(`r printnum(pairs_between_summary$df[2], digits = 1)`) = `r printnum(pairs_between_summary$t.ratio[2], digits = 2)`, *p* < .001, Cohen's *d* = 1.68, 95% CI [1.15, 2.22], representing a large effect. This between-group difference increased further at 3-month follow-up, *t*(`r printnum(pairs_between_summary$df[3], digits = 1)`) = `r printnum(pairs_between_summary$t.ratio[3], digits = 2)`, *p* < .001, Cohen's *d* = 2.50, 95% CI [1.88, 3.11], representing a very large effect.

```{r means}
# Create means table
means_table <- data.frame(
  Time = c("Baseline", "Post-treatment", "3-Month Follow-up"),
  ISTDP_M = sprintf("%.2f", emm_summary$emmean[emm_summary$Tx_condition == "ISTDP"]),
  ISTDP_SE = sprintf("%.2f", emm_summary$SE[emm_summary$Tx_condition == "ISTDP"]),
  ISTDP_CI = sprintf("[%.2f, %.2f]",
                     emm_summary$lower.CL[emm_summary$Tx_condition == "ISTDP"],
                     emm_summary$upper.CL[emm_summary$Tx_condition == "ISTDP"]),
  Waitlist_M = sprintf("%.2f", emm_summary$emmean[emm_summary$Tx_condition == "Control"]),
  Waitlist_SE = sprintf("%.2f", emm_summary$SE[emm_summary$Tx_condition == "Control"]),
  Waitlist_CI = sprintf("[%.2f, %.2f]",
                        emm_summary$lower.CL[emm_summary$Tx_condition == "Control"],
                        emm_summary$upper.CL[emm_summary$Tx_condition == "Control"])
)

apa_table(
  means_table,
  caption = "Estimated Marginal Means for Depression Scores by Treatment Condition and Time Point",
  note = "Values are estimated marginal means from the linear mixed-effects model with random intercepts. CI = 95\\% confidence interval. ISTDP = Intensive Short-Term Dynamic Psychotherapy.",
  align = c("l", rep("c", 6)),
  col.names = c("Time Point", "$M$", "$SE$", "95\\% CI", "$M$", "$SE$", "95\\% CI"),
  col_spanners = list("ISTDP" = c(2, 4), "Waitlist Control" = c(5, 7))
)
```

### Within-Group Changes

For ISTDP participants, depression scores decreased significantly from baseline to post-treatment (*M*~diff~ = `r printnum(pairs_within_summary$estimate[pairs_within_summary$Tx_condition == "ISTDP" & grepl("Baseline.*Post", pairs_within_summary$contrast)], digits = 2)`, *SE* = `r printnum(pairs_within_summary$SE[pairs_within_summary$Tx_condition == "ISTDP" & grepl("Baseline.*Post", pairs_within_summary$contrast)], digits = 2)`, *p* < .001) and continued to decrease from baseline to 3-month follow-up (*M*~diff~ = `r printnum(pairs_within_summary$estimate[pairs_within_summary$Tx_condition == "ISTDP" & grepl("Baseline.*Follow", pairs_within_summary$contrast)], digits = 2)`, *SE* = `r printnum(pairs_within_summary$SE[pairs_within_summary$Tx_condition == "ISTDP" & grepl("Baseline.*Follow", pairs_within_summary$contrast)], digits = 2)`, *p* < .001). Additional improvement occurred between post-treatment and follow-up (*M*~diff~ = `r printnum(pairs_within_summary$estimate[pairs_within_summary$Tx_condition == "ISTDP" & grepl("Post.*Follow", pairs_within_summary$contrast)], digits = 2)`, *SE* = `r printnum(pairs_within_summary$SE[pairs_within_summary$Tx_condition == "ISTDP" & grepl("Post.*Follow", pairs_within_summary$contrast)], digits = 2)`, *p* < .001).

In contrast, waitlist control participants showed minimal change in depression scores across all time points (all *p*s > .10), with no significant differences between baseline and post-treatment (*M*~diff~ = `r printnum(pairs_within_summary$estimate[pairs_within_summary$Tx_condition == "Control" & grepl("Baseline.*Post", pairs_within_summary$contrast)], digits = 2)`, *p* = `r printp(pairs_within_summary$p.value[pairs_within_summary$Tx_condition == "Control" & grepl("Baseline.*Post", pairs_within_summary$contrast)])`), baseline and follow-up (*M*~diff~ = `r printnum(pairs_within_summary$estimate[pairs_within_summary$Tx_condition == "Control" & grepl("Baseline.*Follow", pairs_within_summary$contrast)], digits = 2)`, *p* = `r printp(pairs_within_summary$p.value[pairs_within_summary$Tx_condition == "Control" & grepl("Baseline.*Follow", pairs_within_summary$contrast)])`), or post-treatment and follow-up (*M*~diff~ = `r printnum(pairs_within_summary$estimate[pairs_within_summary$Tx_condition == "Control" & grepl("Post.*Follow", pairs_within_summary$contrast)], digits = 2)`, *p* = `r printp(pairs_within_summary$p.value[pairs_within_summary$Tx_condition == "Control" & grepl("Post.*Follow", pairs_within_summary$contrast)])`).

### Effect Sizes

```{r effect-sizes, fig.cap = "Between-group effect sizes (Cohen's d) over time with 95\\% confidence intervals. Positive values indicate lower depression in the ISTDP group compared to waitlist control. The dashed line represents the threshold for a large effect (d = 0.80).", fig.width = 8, fig.height = 6}
# Calculate effect sizes at each time point
effect_sizes <- data.frame()

for (time_point in c("Baseline", "Post-treatment", "Follow-up")) {
  df_time <- data_clean %>% filter(Time == time_point)

  control_scores <- df_time %>% filter(Tx_condition == "Control") %>% pull(Depression)
  istdp_scores <- df_time %>% filter(Tx_condition == "ISTDP") %>% pull(Depression)

  d_result <- cohen.d(istdp_scores, control_scores)

  cohens_d <- -d_result$estimate
  ci_lower <- -d_result$conf.int[2]
  ci_upper <- -d_result$conf.int[1]

  time_numeric <- ifelse(time_point == "Baseline", 0,
                        ifelse(time_point == "Post-treatment", 1, 2))

  effect_sizes <- rbind(effect_sizes, data.frame(
    Time = time_point,
    Time_numeric = time_numeric,
    Cohens_d = cohens_d,
    CI_lower = ci_lower,
    CI_upper = ci_upper
  ))
}

effect_sizes$Time <- factor(effect_sizes$Time, levels = c("Baseline", "Post-treatment", "Follow-up"))

# Create effect size plot
ggplot(effect_sizes, aes(x = Time_numeric, y = Cohens_d)) +
  geom_hline(yintercept = 0, linetype = "solid", color = "gray40", size = 0.8) +
  geom_hline(yintercept = 0.8, linetype = "dashed", color = "gray50", size = 0.8) +
  geom_ribbon(aes(ymin = CI_lower, ymax = CI_upper, group = 1), alpha = 0.3, fill = "#E63946") +
  geom_errorbar(aes(ymin = CI_lower, ymax = CI_upper), width = 0.1, color = "#E63946", size = 1) +
  geom_line(aes(group = 1), color = "#E63946", size = 1.5) +
  geom_point(color = "#E63946", size = 5, shape = 21, fill = "white", stroke = 2) +
  geom_label(aes(label = sprintf("d = %.2f", Cohens_d)),
            nudge_y = 0.25, fontface = "bold") +
  annotate("text", x = 2.2, y = 0.8, label = "Large effect",
           size = 3, hjust = 0) +
  scale_x_continuous(breaks = c(0, 1, 2),
                    labels = c("Baseline", "Post-treatment", "3-Month\nFollow-up"),
                    limits = c(-0.2, 2.5)) +
  coord_cartesian(ylim = c(-0.5, 3.0), clip = "off") +
  labs(
    x = "Time Point",
    y = "Cohen's d (ISTDP vs. Waitlist Control)"
  ) +
  theme_apa() +
  theme(plot.margin = margin(10, 60, 10, 10))
```

Figure \@ref(fig:effect-sizes) displays the between-group effect sizes (Cohen's *d*) at each time point. Effect sizes increased substantially over time, from negligible at baseline (*d* = `r printnum(effect_sizes$Cohens_d[1], digits = 2)`, 95% CI [`r printnum(effect_sizes$CI_lower[1], digits = 2)`, `r printnum(effect_sizes$CI_upper[1], digits = 2)`]) to large at post-treatment (*d* = `r printnum(effect_sizes$Cohens_d[2], digits = 2)`, 95% CI [`r printnum(effect_sizes$CI_lower[2], digits = 2)`, `r printnum(effect_sizes$CI_upper[2], digits = 2)`]) and very large at 3-month follow-up (*d* = `r printnum(effect_sizes$Cohens_d[3], digits = 2)`, 95% CI [`r printnum(effect_sizes$CI_lower[3], digits = 2)`, `r printnum(effect_sizes$CI_upper[3], digits = 2)`]).

## Model Diagnostics

The random intercept variance was `r printnum(as.numeric(VarCorr(model_final)$id[1]), digits = 2)`, indicating substantial between-person variability in baseline depression levels. The residual variance was `r printnum(sigma(model_final)^2, digits = 2)`. Visual inspection of residual plots (not shown) indicated acceptable model fit, with approximately normally distributed residuals and homogeneous variance across predicted values.

\newpage

# References

::: {#refs custom-style="Bibliography"}
:::

\newpage

# Supplementary Materials

The complete dataset and analysis code are publicly available on the Open Science Framework at https://doi.org/10.17605/OSF.IO/75PU8.
